---
title: "t-SNE algorithm with population genetics data"
author: "Pietro Delfino"
date: "`r format (Sys.Date())`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rtsne); library(tidyverse); library(plotly);
library(poppr); library(data.table)
```

## The data
The dataset here is a table displaying microsatellite size info for 214 grapevine cultivars, genotyped at 45 loci. The info is stored in the **Genalex** format as explained in the **poppr** package vignette. In the first row we have loci number, number of individuals, number of probable populations (as obtained in this case by **strauto** software [http://www.crypticlineage.net/pages/software.html]), and the number of individuals belonging to each one of the population. Populations in this case are called with letters from *a* to *f*. There are missing values as empty cells

## My IDEA
My goal is to apply t-SNE algorithm to SSR data without *a priori* knowledge on populations subdivisions, to see if this technique is able to identify the subgroups as identified with other methods such as Bayesian methods developed with the software **structure** (https://web.stanford.edu/group/pritchardlab/structure.html). The groups in the original file have indeed been identified with the above mentioned software. I will compare the two classifications to see if they are consistent

## Load and explore the data
I will first load the dataset as normal csv file to show the formatting, then I'll load it again but using the **poppr** function to read Genalex formatted files

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

geno <- read_csv("data/genotypes.csv",
                 col_names = FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
DT::datatable(geno[1:10,1:10], rownames = F, filter = "top")
```

Now load it using specific function to store microsatellite (genetic markers in general) formats

```{r, message=FALSE, warning=FALSE}
geno.poppr <- read.genalex("data/genotypes.csv")

class(geno.poppr)[1]
```

```{r}
geno.poppr
```

So as we said, 214 individuals, 45 loci, 6 subpopulations. We are interested in the attribute `tab` of the `geno.poppr` object, that is storing the info on allele frequencies. Indeed we see that once we load the genalex formatted file, allele size info are converted to allele frequencies (0 = no allele, 1 = 1 allele at that size, 2 = 2 alleles at that size, NA = missing value)

```{r}
geno.poppr$tab[1:10, 1:4]

dim(geno.poppr$tab)
```

We now need to convert the `genclone` object to a `genind` object in order to scale and center the dataset and imputing missing values before running t-SNE algorithm

```{r}
geno.poppr.genind <- genclone2genind(geno.poppr)

# we scale and center the matrix and replace NA values with column means
geno.poppr.genind.imputed <- scaleGen(geno.poppr.genind,
                                      scale = TRUE,
                                      center = TRUE,
                                      NA.method = "mean")
class(geno.poppr.genind.imputed)

geno.poppr.genind.imputed[1:10, 1:4]

dim(geno.poppr.genind.imputed)
```

We are now ready to run t-SNE algorithm using the `Rtsne` package (https://github.com/jkrijthe/Rtsne)

```{r, message=FALSE, warning=FALSE}
set.seed(1234)

tsne <- Rtsne(geno.poppr.genind.imputed,
              dims=2,
              perplexity = 10,
              verbose=T,
              max_iter = 15000)
```

Now for plotting we set some graphical parameters. In particular we give a color to each one of the original subpopulations since we want to see if same colors group together

```{r, dev="svg"}
# exclude first 3 rows since they do not contain subpop info
colors <- rainbow(length(unique(geno[-c(1:3),]$X2)))

names(colors) <- unique(geno[-c(1:3),]$X2)
```

### Basic plot

```{r, dev="svg"}
plot(tsne$Y,
     t = 'n',
     main = "tsne")

text(tsne$Y,
     labels = rownames(geno.poppr.genind.imputed),
     col = colors[geno[-c(1:3),]$X2] )
```

### ggplot2

```{r, dev="svg"}
ggplot(as.data.frame(tsne$Y),
       aes(x = tsne$Y[,1],
           y = tsne$Y[,2],
           col = colors[geno[-c(1:3),]$X2])) +
  geom_point((aes(col = colors[geno[-c(1:3),]$X2])),
             size = 2.5,
             alpha = 0.7,
             show.legend = FALSE) +
  theme_bw() +
  xlab("First dimension") +
  ylab("Second dimension")
```

### Plotly

How many groups do you see? Not biased by colors

```{r, dev="svg", warning=FALSE, message=FALSE}
p <- plot_ly(as.data.frame(tsne$Y),
             x = tsne$Y[,1],
             y = tsne$Y[,2],
             text = rownames(geno.poppr.genind.imputed),
             mode = "markers",
             marker = list(size = 11))
             
p <- layout(p,
            title = "Rtsne on SSR data", 
            xaxis = list(title = "First dimension"),
            yaxis = list(title = "Second dimension"),
            showlegend = FALSE)
p
```

With colors

```{r, dev="svg", warning=FALSE, message=FALSE}
p <- plot_ly(as.data.frame(tsne$Y),
             x = tsne$Y[,1],
             y = tsne$Y[,2],
             text = rownames(geno.poppr.genind.imputed),
             mode = "markers",
             color = colors[geno[-c(1:3),]$X2],
             marker = list(size = 11))
             
p <- layout(p,
            title = "Rtsne on SSR data", 
            xaxis = list(title = "First dimension"),
            yaxis = list(title = "Second dimension"),
            showlegend = FALSE)
p
```

We can see that t-SNE algorithm is able to divide our dataframe matching perfectly the subdivision identified by **structure** software. There is just one individual, "MALVASIA ISTRIANA" (group C), which is, according to t-SNE, closer to group A, and also maybe "TERBASCI_cp2" (group A), which is maybe closer to group B 










